---
title: "Bulk_RNAseq_Analysis"
author: "Neha Chaudhary"
date: "9/29/2022"
output: html_document
---

```{r setup, results = FALSE,warning=FALSE,message=FALSE,echo=TRUE, include=FALSE}
### Step 1) Setup the environment and load necessary libraries
library(knitr)
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	cache = TRUE,
	cache.lazy = FALSE
)
library(tidyverse)
library(RColorBrewer)
library(edgeR)
library(pheatmap)
library(readxl)
library(magrittr)
library(dplyr)
library(ggplot2)
library(data.table)
library(ggfortify)
#library(apeglm)
library(DT)
library(ggrepel)
#library(devEMF)
library(readr)
#library(tximeta)
#library(tximport)
library(SummarizedExperiment)
library(shiny)
library(tibble)
library(clusterProfiler)
library("plotly")
#library(pcaExplorer)
library(EnhancedVolcano)
#library(vsn)
library(limma)
```  


```{r Loading Metadata and count, include=FALSE}
## Step 2) Load Metadata and Count matrix
#counts <- read.delim("/projects/tri-data/IRP/Visium/2022_08_Visium_samples/2022_08_HS_bulk_RNA_seq/Goodreau_s6080_MGI3223_225282_HS/all.gene_counts.tsv")
counts <- read.delim("all.gene_counts.tsv") # Load the gene count matrix from a TSV file
counts <- counts[,c(1,3,8:107)]# Subset specific columns from the counts file
counts <- unique(counts)  # Remove duplicated rows

# Set ensembl id as rownames
rownames(counts) <- counts[,1]  # Use first column (Ensembl id as row names
counts <- counts[,-1]           # Remove the first column after setting row names


# Read metadata file
metadata <- read.xlsx("metadata.xlsx", sheetIndex = 1)  # Load sample metadata from first sheet
metadata$Time.Point[which(metadata$Treatment == "Healthy ")] = "Healthy"
metadata <- as.data.frame(metadata)  # Ensure it is a data.frame
rownames(metadata) <- metadata$Sample_name  # Set rownames of metadata by sample name

# Reorder and subset count data
counts_temp <- counts[,c(2:101)]  # Subset counts for columns 2 to 101 (presumably samples)
counts_temp <- setcolorder(counts_temp, as.character(metadata$Filenames))  # Reorder columns by filenames from metadata
counts <- cbind(counts[,1], counts_temp)  # Combine the original gene symbol column with reordered counts
colnames(counts)[1] <- "gene_symbols"  # Name the first column as gene_symbols
rm(counts_temp)

#For ensembl IDs with same gene symbol, take median value for duplicate genes
# Create mapping between rownames and gene symbols
mapT <- cbind(rownames(counts), counts[,1])  # Map row names to gene symbols

# Create data without the gene_symbols column
data <- counts[,-1]  # Remove gene_symbols column for analysis

# Function: For duplicated gene symbols, merge rows by computing median across samples
mergeF <- function(id, data){
  a <- data[id,]              # Subset data for given IDs (genes)
  a <- apply(a, 2, median)    # Calculate the median for each sample (column)
  return(a)
}

# Apply mergeF to handle duplicate genes, grouping by gene symbol
x <- tapply(mapT[,1], mapT[,2], mergeF, data)
y <- do.call(rbind, x)   # Combine the merged data into one data.frame
y <- as.data.frame(y)


# Prepare the final count matrix
cts <- as.matrix(y)  # Convert the final count data.frame to a matrix
head(cts, 2)  # Display first 2 rows of the final count matrix

# Check sample (column) names consistency
all(colnames(cts) %in% metadata$Filenames)  # Ensure all count matrix columns are present in metadata
all(metadata$Filenames == colnames(cts))    # Test if metadata and counts have the same order
rm(x,y,mapT,data)

# Reordering count matrix columns to match metadata
colnames(cts) <- rownames(metadata)  # Set column names of count matrix to sample names
# Check again whether the coldata and count matrix are ordered the same

#head(cts, 2)  # Show first two rows for confirmation

```


### Step 1) Quality Control

```{r Negative Binomial distribution, echo=FALSE}
#check if the data follows negative binomial distribution(mean < variance)
# Calculate the mean of counts for each gene across selected samples (columns 5 to 18) - same group
mean_counts <- apply(cts[, 5:18], 1, mean)
# Calculate the variance of counts for each gene across the same set of samples
variance_counts <- apply(cts[, 5:18], 1, var)
df <- data.frame(mean_counts, variance_counts)  # Create a data frame with mean and variance columns for plotting

# Plot the relationship between mean and variance
# Points: Each point represents a gene's mean and variance across samples
# Line: The red line (variance = mean) is drawn for reference; points above this suggest overdispersion (variance > mean), which is characteristic of a negative binomial distribution
# Log scales: Both axes are log-transformed to better visualize wide ranges of mean/variance values

ggplot(df) +
  geom_point(aes(x=mean_counts, y=variance_counts)) + 
  geom_line(aes(x=mean_counts, y=mean_counts, color="red")) +
  scale_y_log10() +
  scale_x_log10()
 
rm(df, mean_counts, variance_counts)
```

Figure:  Negative Binomial distribution  
Incase of Binomial distribution the variance is greater than the mean and hence the curve. The straight red line indicated mean == variance which is the condition for Poisson distribution. 

### Step 2) Filtering and Differential gene expression analysis (limma)

Number of genes pre filtering:  

```{r filter and limma, echo=FALSE}
## Create DESeq2Dataset object
# Create grouping factor based on metadata
groups <- factor(metadata$Combine)  # Define sample groups for differential analysis

# Create DGEList object using counts and group info
dge <- DGEList(counts=cts, group=groups)  # DGEList is used for edgeR differential analysis

# Filter out genes with low read counts

dim(dge)[1]  # Check the number of genes before filtering

# Calculate Counts Per Million (CPM) normalized by library size (Jing's method for proteomics)
m <- sweep(dge$counts, 2, 1e6 / dge$samples$lib.size, `*`)  # Normalize counts for each sample

min_count_no <- 0.2 * length(groups)  # Minimum number of samples required for each gene (20% of samples)
ridx <- rowSums(m > 2) >= floor(min_count_no)  # Identify genes with CPM >2 in enough samples

# Keep only filtered genes in the DGEList object
dge <- dge[ridx,]  # Subset DGEList to keep robustly expressed genes

# Check the number of genes/features remaining in the DGEList after filtering
# dim(dge)[1] returns the number of rows (genes) in your filtered DGEList

dim(dge)[1]  # Number of genes/features after filtering
rm(m, min_count_no,ridx)
```

Number of genes post filtering:  

```{r limma2, echo=FALSE}
# Normalize library sizes using the TMM method, which adjusts for compositional bias
# This prepares your count data for more accurate downstream comparisons

dge <- calcNormFactors(dge, method = "TMM")  # Calculates normalization factors using TMM and adds them to DGEList

# Set up the design matrix for differential expression analysis
# The design matrix encodes the experimental grouping of samples for model fitting
nf <- calcNormFactors(dge$counts,method="TMM")
design <- model.matrix(~0 + dge$samples$group, data = dge$samples) # Builds a design matrix (no intercept) for group comparison

# Estimate dispersion of counts, which reflects sample-to-sample or biological variability
dge <- estimateCommonDisp(dge, verbose = TRUE)  # Estimate common dispersion across all genes
bcv <- sqrt(dge$common.dispersion)  # Calculate the biological coefficient of variation

# Run voom transformation to convert counts into log2-counts per million (logCPM) with associated precision weights
# voom adjusts for mean-variance relationship, making data suitable for linear modeling (limma)
# The plot argument will produce a mean-variance trend plot
# lib.size ensures library size normalization using precomputed factors
vm <- voom(dge$counts,design,plot=TRUE,
           lib.size=colSums(dge$counts)*nf, save.plot=T)
rm(bcv)
```  



```{r PLot count data, echo=FALSE,fig.height=6, fig.width=8}
# Visualize raw counts for each sample
# Summing raw counts for each sample before normalization
# 'cts' is your raw count matrix; 'colSums(cts)' computes total reads per sample
# 'metadata$Sample_name' assigns sample names to each column
# 'metadata$Combine' assigns each sample to its group

df_dds_preN <- data.frame(
  counts = colSums(cts),                # Total counts (reads) per sample before normalization
  sample = metadata$Sample_name,         # Sample identifiers
  group = metadata$Combine               # Group label for each sample
)

# Visualize raw counts for each sample
# Each bar represents total counts for a sample, colored by group
# x-axis: sample, y-axis: total counts
# Rotating axis text aids readability for many samples
# 'theme_bw()' gives a clean background; adjust text angle for clarity

ggplot(df_dds_preN, aes(x = sample, y = counts, fill = group)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  ggtitle("Count before voom normalization")

# Prepare a new data frame for counts after voom normalization
# 'vm$E' contains the normalized expression data (logCPM from voom)
# colSums(vm$E) summarizes normalized expression across all genes, per sample
# Sample/group assignment remains unchanged

df_dds <- data.frame(
  counts = colSums(vm$E),                # Total normalized counts per sample after voom
  sample = metadata$Sample_name,         # Sample identifiers
  group = metadata$Combine               # Group label for each sample
)

# Visualize normalized counts for each sample, using similar settings
# Compare this plot to the pre-normalization plot to observe normalization impact

ggplot(df_dds, aes(x = sample, y = counts, fill = group)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  ggtitle("Count after voom normalization")

rm(df_dds,df_dds_preN)

```  


#### Figure: The plot shows the total number of reads in each sample - pre and post normalization

```{r Week13_healthy_PCA, echo=FALSE, warning=FALSE,fig.height=7, fig.width=10}
# Get normalized expression data (log2-CPM from voom)
Normalized_data <- vm$E # 'vm$E' contains normalized expression values from voom()

# Subset metadata for samples at Week 13 or Healthy control
meta_sub <- metadata[metadata$Time.Point == "Week 13"|metadata$Time.Point == "Healthy",]
# Subset data matrix to include only relevant samples
data1 <- as.data.frame(Normalized_data[,meta_sub$Sample_name])

# Calculate SD for each gene across selected samples
l <- apply(data1,1,sd)
# Sort genes by SD in decreasing order (most variable genes first)
l <- sort(l, decreasing = T)
# Select the top 1000 most variable genes
d <- data1[names(l)[1:1000],]

# Perform PCA on the selected genes (samples as observations)
pca_ori <- prcomp(t(d), scale. = TRUE)  # Transpose so columns are genes
summary(pca_ori)  # Summary of PCA variance explained


# Prepare data frame for ggplot
# principal components + sample annotations
df_out <- as.data.frame(pca_ori$x)
df_out$Treatment <- meta_sub$Treatment
df_out$Group <- meta_sub$Combine
df_out$response <- meta_sub$Response
df_out$timepoint <- meta_sub$Time.Point
df_out$condition <- meta_sub$condition

# Calculate percentage of variance explained by each principal component
PoV <- pca_ori$sdev^2 / sum(pca_ori$sdev^2)  # Proportion of variance
Percentage_var <- PoV * 100

# Build the PCA scatter plot
ggplot(df_out, aes(x=PC1, y=PC2, color=Group)) +
  xlab(paste("PC1(", round(Percentage_var[1], 2), "%)", sep="")) +
  ylab(paste("PC2(", round(Percentage_var[2], 2), "%)", sep="")) +
  scale_shape_manual(values=c(3,4,7,8,15,16,17,18,10)) +
  geom_point(size=3) +
  stat_ellipse()  # Add ellipses for groups
```




```{r Week13 NR pCA plot}
# Subset metadata for Week 13 non-responders
meta_sub <- metadata[metadata$Response == "NR",]  # Keep only samples designated 'NR' (non-responders)
meta_sub <- meta_sub[meta_sub$Time.Point == "Week 13",]  # Further subset to only those at 'Week 13'

# Extract expression data for these samples
data1 <- as.data.frame(Normalized_data[,meta_sub$Sample_name])

# Calculate standard deviation for each gene across selected samples
l <- apply(data1, 1, sd)  # Compute SD gene-wise
l <- sort(l, decreasing = TRUE)  # Sort genes by variability

d <- data1[names(l)[1:1000], ]  # Select top 1000 most variable genes
library(ggfortify)  # For advanced plotting support

# Perform PCA on transposed data (samples as rows)
pca_ori <- prcomp(t(d), scale. = TRUE)
#summary(pca_ori)  # Shows proportion of variance explained and other stats

# Prepare data frame for plotting
# Combine principal components with sample annotations
df_out <- as.data.frame(pca_ori$x)
df_out$Treatment <- meta_sub$Treatment  # Add treatment info
df_out$response <- meta_sub$Response  # Add response status
df_out$Group <- meta_sub$Combine      # Add combined group label

# Calculate percentage of variance explained by PC1 and PC2
PoV <- pca_ori$sdev^2 / sum(pca_ori$sdev^2)  # Eigenvalues normalized
Percentage_var <- PoV * 100

# Build the PCA plot
# Color points by 'Group'; customize axis labels with variance explained
ggplot(df_out, aes(x=PC1, y=PC2, color=Group)) +
  xlab(paste("PC1(", round(Percentage_var[1], 2), "%)", sep="")) +
  ylab(paste("PC2(", round(Percentage_var[2], 2), "%)", sep="")) +
  scale_shape_manual(values=c(3,4,7,8,15,16,17,18,10)) + # Manual shape mapping if shape aes used
  geom_point(size=3) +      # Plot points, size 3
  stat_ellipse()            # Add group ellipses to visualize spread

```



```{r deg between comparisons}
# Create grouping factor from 'Combine' column in metadata (after cleaning names)
groups <- factor(make.names(metadata$Combine))

# Build design matrix for limma analysis using the sample groups
design <- model.matrix(~0 + groups, data = dge$samples)

# Define contrasts of interest for the analysis
contrast.mat <- makeContrasts(
  ADA.vs.Placebo_NR_Week13 = groupsADA_NR_Week.13 - groupsPlacebo_NR_Week.13,
  ADA.vs.Placebo_R_Week13 = groupsADA_R_Week.13 - groupsPlacebo_R_Week.13,
  levels = design
)

# Fit linear model to normalized expression values
fit <- lmFit(vm, design)
# Apply contrasts to fitted model
fit2 <- contrasts.fit(fit, contrast.mat)
# Empirical Bayes moderation of statistics
fit2 <- eBayes(fit2)

# Optionally remove variables not needed from the environment (except important ones)
# rm(list = setdiff(ls(), c('metadata', 'Normalized_data', 'fit2', 'vm')))

i = i - 1  # Decrement index (likely for iterating through contrasts)
# Get all significant differentially expressed genes for the selected contrast
DEG <- topTable(
  fit2,
  coef = colnames(fit2$coefficients)[i],
  number = nrow(vm$E),
  adjust.method = 'fdr'
)
# Add gene symbols as the first column
DEG <- cbind(rownames(DEG), DEG)
colnames(DEG)[1] <- 'Genes'
# Build output filename for results (using contrast name)
filename <- paste(colnames(fit2$coefficients)[i], '.csv', sep = '')
# filename now contains the name of the contrast (e.g., 'ADA.vs.Placebo_NR_Week13.csv')

```

