---
title: "GSEA_loop_HS_Visium"
author: "Neha Chaudhary"
date: "10/11/2023"
output: html_document
---


### Step 1) Setup the environment and load necessary libraries

```{r setup, results = FALSE,warning=FALSE,message=FALSE,echo=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = FALSE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE,
	cache.lazy = FALSE
)
#BiocManager::install("clusterProfiler")
#BiocManager::install("pathview")
#BiocManager::install("enrichplot")
library(clusterProfiler)
library(enrichplot)
# we use ggplot2 to add x axis labels (ex: ridgeplot)
library(ggplot2)
#BiocManager::install("org.Mm.eg.db")
library(org.Hs.eg.db)
#set path
setwd("GSEA/")
#create folder to save the results
if (!dir.exists("GSEA_results_tables/")){
dir.create("GSEA_results_tables/")
} else {
    print("Dir already exists!")
}

```


### Step 2) Data Preprocesing

Gene set enrichment analysis is run using the GO Biological processes database.

Pathways with Adjusted P value <=0.05 were saved to file.

For the figures, any pathway involved in metabolism or reproduction (contains "repro" or "metabol" in description) have been excluded but are included in the tables.  


```{r preprocessing}

# Assign path to the differential expression gene (DEG) file
file1 <- "path to deg"

# Read in the DEG file as a table, setting the first column as row names and specifying comma separator
deg <- read.delim(file1, header=T, sep = ",", row.names = 1)

# Extract log2 fold change values for each gene
original_gene_list <- deg$logFC

# Set gene names as the names of the vector
names(original_gene_list) <- rownames(deg)

# Remove any NA values (missing data) from the gene list
gene_list <- na.omit(original_gene_list)

# Sort the gene list in decreasing order, as required by clusterProfiler
gene_list = sort(gene_list, decreasing = TRUE)

# Set a random seed for reproducibility
set.seed(123456)

# Run Gene Set Enrichment Analysis (GSEA) using clusterProfiler's gseGO function
#  - geneList: ordered vector of gene logFC values
#  - ont: ontology, "BP" for Biological Process
#  - OrgDb: organism database, "org.Hs.eg.db" for human
#  - keyType: gene identifier type, here SYMBOL
#  - pvalueCutoff: p-value threshold, set to 1 for all results
gse <- gseGO(geneList=gene_list, 
             ont = "BP",
             OrgDb = org.Hs.eg.db,
             keyType = "SYMBOL",
             pvalueCutoff = 1)

# Extract the results from the gseGO output
gsea_results <- gse@result

# Construct output file name by appending the input file name (minus ".csv")
file <- paste("GSEA_results_tables/GSEA_GO_BP_", gsub(".csv","",file), ".txt", sep = "")

# Save the results table to a tab-delimited text file, without quotes
write.table(gsea_results, file, quote=F, sep="\t")

```


### Dotpot

Number of significantly enriched pathways - `r dim(gsea_results)[1]`

Number of pathway shown below ~ 30

```{r dotplot, fig.height=10, fig.width=8}

#dotplot
require(DOSE)
dotplot(gse, showCategory= 15
        , split=".sign") + facet_grid(.~.sign) + theme(axis.text.y = element_text(size = 8))

```



